---
title: "Are differentially methylated regions within genes associated with mantle cell lymphoma?"
author: "Pascal Lafrenz, Mari Hambardzumyan, Lea Herzel, Franziska Lam"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---
```{r, include=FALSE}
library(kableExtra)
```
The alterations in cancer cell DNA methylation pattern have long been detected (Vogelstein and Feinberg, 1983). In many cancer types methylation level has also been shown to have effects on gene deregulation,  (Melnick, 2013). But not much research has been done associated with methylation in *mantle cell lymphoma* (MCL) cells. Therefore in the following project we want to address the following question.

# Research question

Are differentially methylated regions within genes associated with mantle cell lymphoma?

For that purpose we received a data set and an annotation file. Below is a short summary of important key points of the data set.

# Introduction

MCL is a B cell lymphoma which has its origin in the inner mantle zone of the lymph nodes (Hiddemann and Bartram, 2010). The most characteristic mutation in MCL is the translocation between the long arm of chromosome 11 and the long arm of chromosome 14 (t(11;14)(q13;q32)) (Hiddemann and Bartram, 2010). This mutation affects the heavy chains of immunoglobuline and leads to a cycline D1 (CCD1) overexpression (Hiddemann and Bartram, 2010). Cyclin D1 is a transcription factor which controlls the cell cycle (Hiddemann and Bartram, 2010). Through the overexpression of CCD1 the cell is able to proliferate but it can also influence the gene expression of other genes (Hiddemann and Bartram, 2010). It is observed, that a CCD1 overexpression leads to a deregulation of transcription factors and repair genes (Hiddemann and Bartram, 2010). 

Our data from MCL patients and the healty control group originate from a project called BLUEPRINT (Blueprint Project). It was a research association from 2011 to 2016 with the goal to understand the geneactivation and generexpression better (Blueprint Project). To do so they generated referecne epigenomes (Blueprint Project). They focused on certain haematopoietic cells from healthy and leukeamic people, but also on some auto immune diseases (Blueprint Project). 

Epigenetic modifications have an impact on DNA expression (Gibney and Nolan, 2010) and are important in some tumor genesis (Horn et al., 2009). These modifications are for example DNA methylation, histone modification and RNA mediated processes (Gibney and Nolan, 2010). In our project we want to focus on DNA methylation. In humans only cytosine can be methylated to 5-mehtylcytosine (5mC) (Walter and Hümpel, 2017). If the methylation should be inherited it needs to be in a CpG dimer, because in such a dimer there is a cytosine in the mother and the doughter strand (Horn et al., 2009). After the DNA replication the enzyme DNA-methyltransferase 1 is able to methylate the doughter stand if there is a methypation in the CpG of the mother strand (Horn et al., 2009). During the lifetime of a cell different genes need to be expressed and therefore methylated. DNA methyltransferase 3a and 3b is able to methylate cytosine even though the there is no mehtylation in the complementary strand (Horn et al., 2009).
Methylation in promoters lead normally to an inactivation of the the following gene (Horn et al., 2009). But there is also the possibility of activating a gene by methylation. For this the methylation needs to be in the gene.

Quelle??

For an analysis of DNA methylation a few more steps than in a normal DNA-sequencing are required, because it needs to be distinguished between "normal" cytosine and 5-mC. The methylation pattern of the samples in our data set was analyzed by the bisulfite sequencing method (bisulfite seq). The chemical sodium bisulfite converts unmethylated cytosine into uracil (Bock, 2012). The DNA with the converted bases can be amplified by PCR and after that it can be analysed by a next generation sequencing mehtod (Bock, 2012). In our case they used Ilumina. After the analysis of the converted and unconverted DNA both sequences can be compared (Bock, 2012).  If there is a thymine which pairs with a cytosin there was an unmethylated cytosine in this position in the original DNA (Bock, 2012). Thymine which pairs with adenine were originally thymine and every cytosine in the bisulfite treated DNA stands for a 5-mC (Bock, 2012). One problem in the bisulife seq is that bisulfite can not distinguish between 5-mC and 5-hydroxymethylcytosine (5-hmC) (Huang et al., 2010). So we do not really know if every methylated cytosine in our data set is really a 5-mC (Huang et al., 2010). The problem of this fact is that it has been observed that 5-hmC has not the same effect on gene expression as 5-mC has (Moen et al., 2015). Advantages of bisulfite seq are very good reproducibility, a high accuracy and acomprehensive genomic coverage (Bock, 2012).

 


# Data summary

* Bisulfite-Seq data for 10 patients/samples (5 male, 5 female)
* 5 healthy patients, 5 with MCL
* Extracted DNA is from B-cells
* 56175 genes across all samples
* coverage and beta-values for each gene
* Length of each gene

Researching on MCL we found a range of genes which have been linked with this specific cancer, some of which include SOX11, p53, CCND1 etc (Queirós et al., 2016).

To address the research question we broke down the project into 5 parts.

# Project steps



|1. Data processing  | 2. Data normalization and visualization | 3. Data reduction  | 4. Regression and interpretation    | 5. Annotate and interpret results
------------- | ------------- | ------------- | ------------- | -------------
Remove x and y chromosome genes  | Transform beta-values into homoscedastic M-values $$~$$ | Dimensionality reduction, e.g PCA  | Apply t-test        |  
Split data in 4 data frames (cancer vs. control, beta-values and coverage) $$~$$ | Add a column for mean, median and SD for better comparisons  | Apply clustering method and visualize, e.g k-means         | Apply logistic regression        | 
Set a threshold and examine variance for reliable coverage, replace unreliable beta-values with NA $$~$$ | Visualization through plots | 
Separately inspect genes of interest for MCL $$~$$ | Identify batch effect magnitude in the data set
Set a threshold and remove genes under/above threshold, imputate remaining NA's | 

$$~$$

## Part 1: Data processing

For now, we want to single out the gene data frame of the patients to inspect it separately. We will examine only the gene data frame and neglect the data from tiling, promoters and cpg-islands because they are irrelevant to answer our research question.

```{r}
Samples <- readRDS(file = "Mantle-Bcell_list.RDS.gz")
Gene_data_frame <- Samples$genes
dim(Gene_data_frame)
```

There are much information from 56175 genes across all 10 patients. Mostly we are interested in the coverage and the beta values. After the entire data clean-up we still want to retain around 90% of our beta value information (~500000 genes).

We are starting the data clean-up with the sex chromosomes. In our case we have different numbers of female/male patients in healthy/cancer sample groups. In female X chromosome inactivation leads to the silencing of the majority of genes on one X, this would lead to significant differences in the methylation profiles between female/male patients (Sharp et al., 2011). Therefore, we remove gene data of sex chromosomes. 

```{r}
Gene_data_frame_x_y <- 
  Gene_data_frame[-which(Gene_data_frame$Chromosome == "chrX"),]
Gene_data_frame_x_y <- 
  Gene_data_frame_x_y[-which(Gene_data_frame_x_y$Chromosome == "chrY"),]
```

Now we want to go through the coverage values separately to remove any unreliable values, outliers and possible systematic errors. First we separated the coverage values of healthy and cancer patients into 2 data frames. We did the same separation into 2 data frames with the beta-values because we will need these later. All the  other information will be neglectd till the interpretation. 

```{r}
healthy_coverage <- Gene_data_frame_x_y[, 21:25]
cancer_coverage <- Gene_data_frame_x_y[, 26:30]
healthy_beta_values <- Gene_data_frame_x_y[, 11:15]
cancer_beta_values <- Gene_data_frame_x_y[, 16:20]
```

Then we created a histogram with mean coverage values for each gene on a common log scale and added 10 quantiles from 0 % to 100 %.

```{r}
mean_cancer_coverage <- rowMeans(cancer_coverage)
hist(
  log10(mean_cancer_coverage),
  breaks = "fd",
  main = "Cancer coverage: Mean frequency",
  xlab = "Common logarithm of coverages",
  col = "indianred2",
  border = "gray20"
)
abline(v = log10(quantile(
  mean_cancer_coverage,
  probs = seq(0, 1, 0.1),
  na.rm = TRUE
)),
col = "black",
lty = 5,
lwd = 1)


mean_healthy_coverage <- rowMeans(healthy_coverage)
hist(
  log10(mean_healthy_coverage),
  breaks = "fd",
  main = "Healthy coverage: Mean frequency",
  xlab = "Common logarithm of coverages",
    col = "seagreen2",
    border = "gray20"
)
abline(v = log10(quantile(
  mean_healthy_coverage,
  probs = seq(0, 1, 0.1),
  na.rm = TRUE
)),
col = "black",
lty = 5,
lwd = 1)
```

It was essential that we used a common logarithm scale because our coverages were wide spread over a great range and we did not want to seperate the histograms. But it is easy to see that we have two groups of mean coverage values in healthy coverage histogram. In cancer coverage histogram it was harder to decide how to handle the mean values between 10 and 100 coverages. We concluded that we also had two major mean value groups. But overall it is seen that most of our data had a strong coverage value and that we could later adapt our analysis to it. Not even one fifth of the beta values had less than 100 coverages. As previous said, healthy beta values had stronger coverages than cancer beta values. We need to see if this will effect our downstream analysis.

To check how widely spread the coverages across the patients are, we calculated the standard deviations (SD) of each gene across the patients. We plotted the SD-values for cancer and healthy samples on a histogram with a log scale and again quantiles from 0 % to 100 %.

```{r}
sd_cancer_coverage <- apply(cancer_coverage, 1, sd)
hist(log10(sd_cancer_coverage),
     breaks = "fd",
     main = "Cancer coverage: SD frequency",
     xlab = "Common logarithm of coverages",
     col = "indianred2",
     border = "gray20"
)
abline(v = log10(quantile(
  sd_cancer_coverage,
  probs = seq(0, 1, 0.1),
  na.rm = TRUE
)),
col = "black",
lty = 5,
lwd = 1)

sd_healthy_coverage <- apply(healthy_coverage, 1, sd)
hist(log10(sd_healthy_coverage),
     breaks = "fd",
     main = "Healthy coverage: SD frequency",
     xlab = "Common logarithm of coverages",
     col = "seagreen2",
     border = "gray20"
)
abline(v = log10(quantile(
  sd_healthy_coverage,
  probs = seq(0, 1, 0.1),
  na.rm = TRUE
)),
col = "black",
lty = 5,
lwd = 1)
```
Again we can distinguish between two major SD groups in both histograms. It is important to mention, that 90 % of mean cancer coverages SD had way under SD = 10,000. In contrast, the 90 % quantile of mean healthy coverages SD was only slightly under SD = 10,000. A low standard deviation would be optimal but we had to accept our outcome of wide spread standard deviation because most of the data is acceptable spreaded.

By setting the lower coverage threshold to 5 % and the upper to 99.9 %, we fulfilled our initial requirement of retaining around 90 % of the information. This is proven later. 

```{r}
threshold_cancer_lower <-
  quantile(mean_cancer_coverage,
           probs = 0.05,
           na.rm = TRUE)
threshold_cancer_upper <-
  quantile(mean_cancer_coverage,
           probs = 0.999,
           na.rm = TRUE)

threshold_healthy_lower <-
  quantile(mean_healthy_coverage,
           probs = 0.05,
           na.rm = TRUE)
threshold_healthy_upper <-
  quantile(mean_healthy_coverage,
           probs = 0.999,
           na.rm = TRUE)
```

Beta-values with very low or extreme high coverages are not reliable. Therefore with the help of the predefined coverage threshold for each patient we will set the beta values of the genes which didn't fulfill the coverage requirement to NA. We wrote a function which checked every table cell and set unreliable coverage value to zero. After that the associated table cells in bata values were set to NA.

```{r}
cancer_threshold_function <- function(cancer_coverage) {
  if(cancer_coverage <= threshold_cancer_lower) {
    return("NA")}
  else {return(cancer_coverage)}
  
  if(cancer_coverage >= threshold_cancer_upper) {
    return("NA")}
  else{return(cancer_coverage)}
}

cancer_coverage <- apply(cancer_coverage, MARGIN = c(1,2), FUN = cancer_threshold_function)

cancer_coverage[cancer_coverage == "NA"] <- NA 
cancer_beta_values[cancer_coverage == "NA"] <- NA


healthy_threshold_function <- function(healthy_coverage) {
  if(healthy_coverage <= threshold_healthy_lower) {
    return("NA")}
  else {return(healthy_coverage)}
  
  if(healthy_coverage >= threshold_healthy_upper) {
    return("NA")}
  else{return(healthy_coverage)}
}

healthy_coverage <- apply(healthy_coverage, MARGIN = c(1,2), FUN = healthy_threshold_function)


healthy_coverage[healthy_coverage == "NA"] <- NA
healthy_beta_values[healthy_coverage == "NA"] <- NA
```

Now we needed to deal with NA's in the beta-values, because they represent no information and they could cause errors in our further analysis. We had many NA's but we cannot remove every gene with a NA because this would lead to a loss of too much information.
Our sum of NA's is:
```{r}
sum(is.na(cancer_beta_values))
sum(is.na(healthy_beta_values))
```

We decided that we want to remove each gene with 3 or more NA's in 'healthy_beta_values' or 'cancer_beta_values', because if there are not enough values our results will be not meaningful. To do so we added two new columns to each data frame. One column is for the number of NA's per gene in healthy samples and one for the number of NA's per gene in cancer samples. We need both columns in both data frames because it makes it easier to remove the rows with too much NA's later.

```{r}
cancer_beta_values$Number_of_NA_cancer <-
  rowSums(is.na(cancer_beta_values))
cancer_beta_values$Number_of_NA_healthy <-
  rowSums(is.na(healthy_beta_values))

healthy_beta_values$Number_of_NA_healthy <-
  rowSums(is.na(healthy_beta_values))
healthy_beta_values$Number_of_NA_cancer <-
  rowSums(is.na(cancer_beta_values))
```

To visualize the number of NA's in the beta-values we plotted different histograms.

```{r, include=FALSE}

library(tidyverse)
library(gridExtra)
library(ggrepel)
library(plotly)

```
```{r error=FALSE, warning=FALSE}

p1 <- ggplot() +
  geom_bar(
    data = cancer_beta_values,
    mapping = aes(x = cancer_beta_values$Number_of_NA_cancer),
    fill = "indianred2"
  ) +
  ggtitle("NA's per Gene in \n cancer samples") +
  labs(x = "Number of NA's", y = "Number of Genes") +
  theme(plot.title = element_text(
    color = "black",
    size = 14,
    face = "bold",
    hjust = 0.5
  )) 

p2 <- ggplot() +
  geom_bar(
    data = cancer_beta_values,
    mapping = aes(x = cancer_beta_values$Number_of_NA_cancer),
    fill = "indianred2"
  ) +
  xlim(0.5, 5.5) +
  ggtitle("NA's per Gene in \n cancer samples \n (zoomed in)") +
  labs(x = "Number of NA's", y = "Number of Genes") +
  theme(plot.title = element_text(
    color = "black",
    size = 14,
    face = "bold",
    hjust = 0.5
  )) 

p3 <- ggplot() +
  geom_bar(
    data = healthy_beta_values,
    mapping = aes(x = healthy_beta_values$Number_of_NA_healthy),
    fill = "seagreen2"
  ) +
  ggtitle("NA's per Gene in \n healthy samples") +
  labs(x = "Number of NA's", y = "Number of Genes") +
  theme(plot.title = element_text(
    color = "black",
    size = 14,
    face = "bold",
    hjust = 0.5
  )) 
  
p4 <- ggplot() +
  geom_bar(
    data = healthy_beta_values,
    mapping = aes(x = healthy_beta_values$Number_of_NA_healthy),
    fill = "seagreen2"
  ) +
  xlim(0.5, 5.5) +
  ggtitle("NA's per Gene in \n healthy samples \n (zoomed in)") +
  labs(x = "Number of NA's", y = "Number of Genes") +
  theme(plot.title = element_text(
    color = "black",
    size = 14,
    face = "bold",
    hjust = 0.5
  )) 
grid.arrange(p1, p2, ncol = 2)
grid.arrange(p3, p4, ncol = 2)
```

The first histogram of cancer/healthy is to get an overview over all numbers of missing values per gene. Most genes in 'cancer_beta_values' and 'healthy_beta_values' have no NA's. That is good, because it means we have much information in our analysis and our results will be more meaningful.
We wanted to see the differences between 2, 3, 4 and 5 NA's per gene better. To do so we reduced the range of the axis. Now it is visible that in 'healty_beta_values' a little amount of genes have 5 NA's. In comparison to that 'cancer_beta_values' have more genes with 5 NA's but less genes with 2, 3 or 4 NA's.


Then we removed the genes with 3 or more missing values. Here it was important to remove the genes in both data frames, because a gene in only one data frame without values to compare is meaningless. After the deletion we checked if the rownames are the same in both data frames and it was the case. This is important because otherwise the wrong rows would be compared later and the results would be completely wrong. 

```{r}
cancer_beta_values <-
  cancer_beta_values[-which(
    cancer_beta_values$Number_of_NA_cancer >= 3 |
      cancer_beta_values$Number_of_NA_healthy >= 3
  ),]
healthy_beta_values <-
  healthy_beta_values[-which(
    healthy_beta_values$Number_of_NA_cancer >= 3 |
      healthy_beta_values$Number_of_NA_healthy >= 3
  ),]
sum(rownames(healthy_beta_values) != rownames(cancer_beta_values))
```

We removed the columns with the NA's-values because we did not need them anymore. 

```{r}
cancer_beta_values <-
  cancer_beta_values[,-which(
    colnames(cancer_beta_values) %in% c("Number_of_NA_cancer", "Number_of_NA_healthy")
  )]
healthy_beta_values <-
  healthy_beta_values[,-which(
    colnames(healthy_beta_values)  %in%  c("Number_of_NA_cancer", "Number_of_NA_healthy")
  )]

```

We can not keep the remaining NA's in the beta-values because this will cause errors in our downstream analysis. But we also did not want to delete the genes, so we decided to replace the NA's with the mean value of the respective gene.
To cut things short and to make it a little easier for our next analysis, we transposed both beta value data frames (cancer and healthy). Afterwards we transposed both data frames back for further analysis.

```{r}
transposed_cancer_beta_values <- t(cancer_beta_values)
transposed_healthy_beta_values <- t(healthy_beta_values)

for (i in 1:ncol(transposed_cancer_beta_values)) {
  transposed_cancer_beta_values[is.na(transposed_cancer_beta_values[, i]), i] <-
    mean(transposed_cancer_beta_values[, i], na.rm = TRUE)
}
for (i in 1:ncol(transposed_healthy_beta_values)) {
  transposed_healthy_beta_values[is.na(transposed_healthy_beta_values[, i]), i] <-
    mean(transposed_healthy_beta_values[, i], na.rm = TRUE)
}

cancer_beta_values <- data.frame(t(transposed_cancer_beta_values))
healthy_beta_values <- data.frame(t(transposed_healthy_beta_values))
```

We proofed that there are no more NA's left in the cancer and healthy data frames.

```{r}
sum(is.na(transposed_cancer_beta_values))
sum(is.na(transposed_healthy_beta_values))
```

There are some important genes for MCL which have ofen a different methylation pattern in comparison to heatlhy b-cells. These genes are:

Gene   |Ensembl         |Function
:---------:|:----------------------:|:-------------------------------------------------------------------:|
CCND1  |ENSG00000110092 | Cycline D1 controlls cell cycle (Fu et al., 2004). CCND1 is essential for the maintenance of established MCL tumor cell lines and preserves genomic stability during DNA replication in MCL cells (Mohanty et al., 2017).  
SOX11	 |ENSG00000176887 | SRY-box related gene	(SRY = sex determining region Y chromosome ) (Amann-Vesti et al., 2018). Critical for survival of neuronal and mesenchymal progenitor cells (Mohanty et al., 2019). One reason for expression in MCL is CCND1 (Mohanty et al., 2019). Epigenetically regulated (Wasik et al., 2013). Promotes oncogenic proliferation (Kuo et al., 2018).
NR2F2	 |ENSG00000185551	| Is a nuclear receptor (https://www.ncbi.nlm.nih.gov/gene/7026, 6-6-2019) and plays a role as transcription factor in the differentiation and pluripotency of human embryonic stem cells (Rosa and Brivanlou, 2011).
p53  	 |ENSG00000141510	| Transcription factor which can trigger a variety of antiproliferative programs by activating or repressing key effector genes. Modulates e.g. apoptosis, cell-cycle arrest, senescence, or modulation of autophagy (Zilfou and Lowe, 2009). 
AHR  	 |ENSG00000106546	|	Activation of aryl hydrocarbon receptor (AhR) results in loss of apoptosis response through overexpression of cyclooxygenase-2 (COX-2) (key role in cancer development - especially in lymphoma). COX-2 upregulates antiapoptotic proteins (Vogel et al., 2007). 
ROBO1	 |ENSG00000169855	| Tumor suppressor gene. Loss of function results in an invasive and metastasic cancer development (Parray et al., 2014). 
SOX9 	 |ENSG00000125398	|SRY-box related gene	(SRY = sex determining region Y chromosome ) (Amann-Vesti et al., 2018). Important in developement of testicle (Amann-Vesti et al., 2018). 
HOXA9	 |ENSG00000078399	|	Belongs to a family of highly conserved transcriptional regulators with critical roles in normal development and hematopoiesis. Effects of HoxA9 overexpression include strong self-renewal and inhibiting differentiation of progenitor cells (Brumatti et al., 2013). 
CDH1 	 |ENSG00000039068	|	Encoding the tumour-suppressor protein E-cadherin (van der Post et al., 2015). Anaphase-promoting complex/cyclosome (APC/C) is dependent on Cdh1. Cdh1 is phosphorylated and inactivated at the Start transition of the new cell cycle (Ondracka et al., 2016). 
CDC14B |ENSG00000081377	|	Important in DNA repair mechanisms and activation of APC/C-CDH1 complex (Zhou et al., 2016). Downregulation leads to genomic instability by enabling cells to down-regulate p53 and override cell cycle checkpoints (Klauck et al., 2018). 
FOXC1	 |ENSG00000054598	| Forkhead box C1 (FOXC1) regulates tumor-associated genes and is regulated by multiple pathways that control its mRNA expression and protein activity (Xu et al., 2016; Yang et al., 2017). Aberrant FOXC1 expression promotes growth, migration, abnormal cell proliferation and angiogenesis therefore functions as an oncogene (Xu et al., 2016). 
G0S2	 |ENSG00000123689	|	Lipolytic inhibitor G0/G1 switch gene 2 (G0S2) plays various important roles in cellular functions such as cell proliferation, apoptosis and oxidative phosphorylation. Upregulation leads to radioresistant cancer (Wang et al., 2019). G0S2 is normally epigenetically silenced (Welch et al., 2009). 
GPX3	 |ENSG00000211445	|	Glutathione peroxidase 3 (GPX3) catalyzes reduction of reactive oxygen species (ROS) in plasma and acts as a redox signal modulator with tumor supressor activity. Protect cells against oxidative damage. Downregulation of GPX3 in many types of human cancer is caused by promoter hypermethylation (An et al., 2018). 
LGALS3 |ENSG00000131981	|	Galectin-3 (LGALS3) elevates inflammatory processes. LGALS3 protects cancer cells by inducing tolerance to drug treatment. Auto-induction of Galectin-3 mRNA and pathways of inflammatory mediators (Fei et al., 2015). 
MAL  	 |ENSG00000172005	|	Myelin and Lymphocyte protein (MAL) is a putative cancer marker associated with chemoresistance and cell survival (Zanotti et al., 2017). 
NPTX2	 |ENSG00000106236	|	 Neuronal pentraxin-2 (NPTX2) is a proinflammatory cytokine. NPTX2 gene is hyper-methylated in MCL and thus non-apoptotic cell death is inhibited (Enjuanes et al., 2013; Li et al., 2018).
PAX6	 |ENSG00000007372	|	Paired box protein (Pax6) is a regulator (transcription factor) of fate determination of precursor cells (Imamura and Greer, 2013). Suppression of PAX6 promotes cell proliferation and inhibits apoptosis (Meng et al., 2014). 
TFPI2	 |ENSG00000105825	|	Tissue factor pathway inhibitor 2 (TFPI2) is a serine protease inhibitor which regulates physiological extracellular matrix remodeling (Puttabyatappa et al., 2017). Abberant methylation (hypermethylation) of TFPI2 gene protects the extracellular matrix of cancer cells from degradation and play an important role in the carcinogenesis of cancer (Hu et al., 2017). TFPI2 is identified as a tumor suppressor (Dong et al., 2015). 
THEM4	 |ENSG00000159445	|	Thioesterase Superfamily Member 4 (THEM4) is a mediator which negatively regulates protein kinase B (PKB). PKB is targeted by receptor tyrosine kinasis which signals via phosphatidylinositol 3-kinase (PI3K) . THEM4 is hypermethylated (promoter) in MCL and mediates cell survival, growth, proliferation, cell migration and angiogenesis (Enjuanes et al., 2011; Rajapakse et al., 2018; Wang et al., 2014; Zhao et al., 2013). 
TWIST1 |ENSG00000122691	|	Twist Family BHLH Transcription Factor 1 (TWIST1) is a key regulator in embryonic development and organogenesis.  TWIST1 is implicated in tumor initiation, stemness, angiogenesis, dissemination, and chemoresistance in various cancer types (Zhao et al., 2017). Overexpression of TWIST1 leads to suppression of transcription of pro-apoptotic BH3-only gene BCL2L11. TWIST1 promoter is hypermethylated in most cancers (Zhao et al., 2017) but has no impact on protein expression. Nevertheless, it can be a biomarker for cancer (Gort et al., 2008).

To be sure we did not removed these genes in our data processing we tested if they are still all in our data frames. Because each of the 2 data frames contains the same genes it is enough to check if they are part of one of them.

```{r}
important_genes <-
  data.frame(
    c(
      "ENSG00000176887",
      "ENSG00000185551",
      "ENSG00000141510",
      "ENSG00000110092",
      "ENSG00000106546",
      "ENSG00000169855",
      "ENSG00000125398",
      "ENSG00000078399",
      "ENSG00000039068",
      "ENSG00000081377",
      "ENSG00000054598",
      "ENSG00000123689",
      "ENSG00000211445",
      "ENSG00000131981",
      "ENSG00000172005",
      "ENSG00000106236",
      "ENSG00000007372",
      "ENSG00000105825",
      "ENSG00000159445",
      "ENSG00000122691"
    )
  )
important_genes_in_data_set <- data.frame()
for (i in 1:nrow(important_genes)) {
  important_genes_in_data_set[i, ] <-
    cancer_beta_values[which(row.names(cancer_beta_values) == important_genes[i, ]), ]
}
sum(is.na(important_genes_in_data_set))
```

Now, we can test if we fulfilled our initial requirement of retaining around 90 % of the information.

```{r}
dim(cancer_beta_values)/dim(Gene_data_frame_x_y)
```
It so happens that we kept slightly over 90 % of the information (91.1 %). That is more than we expected and therefore our quality control was successful. 
We can continue to our next step: data normalization and visualization.


$$~$$


## Part 2: Data normalization and vizualisation

In this part we want to normalise the beta values in order to make them more homoscedastic for further statistical tests and downstream analysis. For that purpose we apply logit transformation and turn them into m values. For the conversion to m values we used the following formula: 
$$ M_i~ = log_2(\frac{beta_i}{1-beta_i}) $$

```{r}
cancer_m_values <-
  data.frame(log2(cancer_beta_values / (1 - cancer_beta_values)))
healthy_m_values <-
  data.frame(log2(healthy_beta_values / (1 - healthy_beta_values)))
```

We included mean and standard deviation values to the original data frame, therefore, we can later visualize them.
```{r}
mean_healthy_m_values <- rowMeans(healthy_m_values)
mean_cancer_m_values <- rowMeans(cancer_m_values)
sd_cancer_m_values <- apply(cancer_m_values, 1, sd)
sd_healthy_m_values <- apply(healthy_m_values, 1, sd)
extended_cancer_m_values <-
  cbind.data.frame(cancer_m_values, mean_cancer_m_values, sd_cancer_m_values)
extended_healthy_m_values <-
  cbind.data.frame(healthy_m_values, mean_healthy_m_values, sd_healthy_m_values)

```

Firstly, we want to check if our transformation was successfull by comparing it to the literature.
```{r}
c1 <- ggplot() +
  geom_point(
    mapping = aes(
      x = extended_cancer_m_values$mean_cancer_m_values,
      y = rowMeans(cancer_beta_values)
    ),
    na.rm = TRUE,
    alpha = 1 / 10
  ) +
  labs(x = "Mean cancer beta values",
       y = "Mean cancer m values",
       title = "Comparison of mean values") +
  theme_bw()

c2 <- ggplot() +
  geom_point(
    mapping = aes(
      x = extended_healthy_m_values$mean_healthy_m_values,
      y = rowMeans(healthy_beta_values)
    ),
    na.rm = TRUE,
    alpha = 1 / 10
  ) +
  labs(x = "Mean healthy beta values",
       y = "Mean healthy m values",
       title = "Comparison of mean values") +
  theme_bw()

grid.arrange(c1, c2, ncol = 2)

```

Compared to literature, beta-value to m-value transition worked perfectly well and we can continue our downstream analysis (Du et al., 2010). Aberration of transition occured because of mean value calculation.

Afterwards we changed the endings of the patinets names to make it more clear to which data frame they belonged.

```{r}
colnames(healthy_m_values) <-
  c(
    "Bcell_naive_VB_NBC_NC11_41.M",
    "Bcell_naive_VB_NBC_NC11_83.M",
    "Bcell_naive_VB_S001JP51.M",
    "Bcell_naive_VB_S00DM851.M",
    "Bcell_naive_VB_S01ECGA1.M"
  )

colnames(cancer_m_values) <- c(
  "cancer_VB_S01FE8A1.M",
  "cancer_VB_S01FF6A1.M",
  "cancer_VB_S01FH2A1.M",
  "cancer_VB_S01FJZA1.M",
  "cancer_VB_S01FKXA1.M"
)
```

Thereafter we calculated the means and standard deviations of the m values, subsequently plotting a histogram for each, to see how they are spread.

```{r warning=FALSE}

hist(
  log10(mean_cancer_m_values),
  breaks = "fd",
  main = "Cancer M values: Mean frequency",
  xlab = "Common logarithm of M values",
  col = "indianred2",
  border = "gray20",
  xlim = c(-1, 1)
)
abline(v = log10(quantile(
  mean_cancer_m_values,
  probs = seq(0, 1, 0.1),
  na.rm = TRUE
)),
col = "black",
lty = 5,
lwd = 1)

hist(
  log10(mean_healthy_m_values),
  breaks = "fd",
  main = "Healthy M values: Mean frequency",
  xlab = "Common logarithm of M values",
  col = "seagreen2",
  border = "gray20",
  xlim = c(-1, 1)
)
abline(v = log10(quantile(
  mean_healthy_m_values,
  probs = seq(0, 1, 0.1),
  na.rm = TRUE
)),
col = "black",
lty = 5,
lwd = 1)

hist(
  log10(sd_cancer_m_values),
  breaks = "fd",
  main = "Cancer M values: SD frequency",
  xlab = "Common logarithm of M values",
  col = "indianred2",
  border = "gray20"
)

hist(
  log10(sd_healthy_m_values),
  breaks = "fd",
  col = "seagreen2",
  main = "Healthy M values: SD frequency",
  xlab = "Common logarithm of M values",
  border = "gray20"
)
```

As we can see in the mean frequency histograms there are some values of common logarithms of m values which are more than 1.0 and less than -1.0, this is due to the beta values which were registered as 1 or 0 which resulted in m values of +inf and -inf. These will be removed before the principal component analysis. 
Comparing the mean frequency histograms of cancer and healthy samples, we can conclude that both are right-skewed, whereas in healthy m values the effect is more prominent. In the mean m value histogram of healthy patients we can observe that the median lays around 0.6 which corresponds to beta values of around 0.8, in comparison to cancer where the median lays around 0.2 which correspond to beta vlues of 0.6. Important to mention is the fact that  there are far less genes which have low beta values in the healthy samples as in the cancer samples (judging by the tenth quantiles of the mean histograms). 
But overall we can assume a normal distribution in healthy and cancer distribution.

Furthermore, we created a data frame which only include our cancer related genes. This data frame is extended by gene name rather than ensembl name.
```{r}

extended_cancer_m_values_gene <- extended_cancer_m_values[c(
  "ENSG00000176887",
  "ENSG00000185551",
  "ENSG00000141510",
  "ENSG00000110092",
  "ENSG00000106546",
  "ENSG00000169855",
  "ENSG00000125398",
  "ENSG00000078399",
  "ENSG00000039068",
  "ENSG00000081377",
  "ENSG00000054598",
  "ENSG00000123689",
  "ENSG00000211445",
  "ENSG00000131981",
  "ENSG00000172005",
  "ENSG00000106236",
  "ENSG00000007372",
  "ENSG00000105825",
  "ENSG00000159445",
  "ENSG00000122691"
),]

extended_healthy_m_values_gene <- extended_healthy_m_values[c(
  "ENSG00000176887",
  "ENSG00000185551",
  "ENSG00000141510",
  "ENSG00000110092",
  "ENSG00000106546",
  "ENSG00000169855",
  "ENSG00000125398",
  "ENSG00000078399",
  "ENSG00000039068",
  "ENSG00000081377",
  "ENSG00000054598",
  "ENSG00000123689",
  "ENSG00000211445",
  "ENSG00000131981",
  "ENSG00000172005",
  "ENSG00000106236",
  "ENSG00000007372",
  "ENSG00000105825",
  "ENSG00000159445",
  "ENSG00000122691"
),]

extended_cancer_m_values_gene <-
  cbind(
    extended_cancer_m_values_gene,
    Important_genes = c(
      "SOX11",
      "NR2F2",
      "p53",
      "CCND1",
      "AHR",
      "ROBO1",
      "SOX9",
      "HOXA9",
      "CDH1",
      "CDC14B",
      "FOXC1",
      "G0S2",
      "GPX3",
      "LGALS3",
      "MAL",
      "NPTX2",
      "PAX6",
      "TFPI2",
      "THEM4",
      "TWIST1"
    ) 
  )
```

Now, the task was to compare mean cancer m-values with mean healthy m-values. Our goal was, to visualize the differences and similarities in the m-values of cancer and healthy samples. A big resemblance would result in a linear line, while a curve indicates differences in the mean methylation patterns.
$$~$$

The following overview plot includes:

* mean of cancer and healthy m-values (black points), 
* general trend of points to solve overplotting problem (blue smoothed curve),
* straight line through origin (y=x) for better interpreting (blue line)

``` {r}
ggplot() +
  geom_point(
    mapping = aes(
      x = extended_cancer_m_values$mean_cancer_m_values,
      y = extended_healthy_m_values$mean_healthy_m_values
    ),
    na.rm = TRUE,
    alpha = 1 / 10
  ) +
  geom_smooth(
    mapping = aes(
      x = extended_cancer_m_values$mean_cancer_m_values,
      y = extended_healthy_m_values$mean_healthy_m_values
    ),
    na.rm = TRUE,
    alpha = 1 / 10,
    colour = "darkgrey"
  ) +
  labs(x = "Mean cancer m-values",
       y = "Mean healthy m-values",
       title = "Comparison of mean m-values") +
  theme_bw() +
  geom_abline(
    mapping = NULL,
    data = NULL,
    slope = 1,
    intercept = 0,
    colour = "yellow2"
  ) + 
  xlim(-12,9) + 
  ylim(-12,9)

```
We compared mean healthy and cancer m-values. Infinite values are presented at the edges. The general trend indicates that a lot of high methylated genes in healthy samples are not methylated/less methylated in cancer samples. This trend is solidified by the yellow line, shown in the plot, which would indicate a linear trend between cancer and healthy sample methylation. Nevertheless, some genes have gain a higher methylation grade. Overall, it seems that a lot of genes have pretty much the same methylation grade because they are in a huge data cloud around 0 to 5. It would be interesting to see which genes have lost  most of their methylation or gain most of it because it can be an indication of function gain (no methylation) or function loss (methylation increase).

So after that we focused on details to evaluate previous mentioned important genes. The following detailed plot visualizes:

* mean of cancer and healthy m-values (black points), 
* highlights of important genes for cancer (red points), 
* labels of specific genes,
* straight line through origin (y=x) for better interpreting (yellow line)

```{r}
ggplot() +
  geom_point(
    mapping = aes(
      x = extended_cancer_m_values$mean_cancer_m_values,
      y = extended_healthy_m_values$mean_healthy_m_values
    ),
    na.rm = TRUE,
    alpha = 1 / 10
  ) +
  geom_point(
    mapping = aes(
      x = extended_cancer_m_values_gene$mean_cancer_m_values,
      y = extended_healthy_m_values_gene$mean_healthy_m_values
    ),
    colour = "red",
    size = 2
  ) +
  geom_label_repel(
    aes(
      label = Important_genes,
      x = extended_cancer_m_values_gene$mean_cancer_m_values,
      y = extended_healthy_m_values_gene$mean_healthy_m_values
    ),
    data = extended_cancer_m_values_gene,
    point.padding = 0.5,
    label.size = 0.1,
    segment.colour = "cornflowerblue",
    segment.alpha = 0.9
  ) +
  labs(x = "Mean cancer m-values",
       y = "Mean healthy m-values",
       title = "Comparison of mean m-values") +
  theme_bw() +
  xlim(-5, 5) +
  ylim(-5, 5) +
  geom_abline(
    mapping = NULL,
    data = NULL,
    slope = 1,
    intercept = 0,
    colour = "yellow2"
  )
```

For better plot evaluation we draw a line with equation y = x. Thus, we can identify if there is a linear relationship. Overall, our predicted important genes have not fully lost their methylation or gain full methylation. 
As stated in several studies, DNA demethylation in gene regions tends to downregulate gene expression but is also dependent on several key factors, including:

* whether embedded demethylated functional elements regain activity,
* whether demethylated regions are located at intron-exon junctions together with destabilized nucleosomes and
* possible effects on the rates of transcript elongation and splicing. 
(Yang et al., 2014)

Therefore, we can only predict what otucome different methylated gene regions could have.

Gene   | Comparison of methylation between samples | Interpretation
:---------:|:----------------------------:|:-------------------------------------------------------------:|
CCND1  | healthy > cancer | Less expression in cancer samples therefore lower genomic stability during DNA replication in MCL cells.
SOX11  | cancer > healthy | Higher expression in cancer samples therefore SOX11 activation: Change of mature B cells phenotype to aberrant cell type. Promotion of oncogenic proliferation of MCL.
NR2F2  | cancer > healthy | Higher expression in cancer samples thus NR2F2 activation: Gain of pluripotency of MCL.
p53    | healthy > cancer | Less expression in cancer samples thus downregulation: Loss of tumor supressor gene which is important for apoptosis initiation which is therefore disabled in MCL.
AHR    | healthy > cancer | Less expression in cancer samples thus downregulation: No stabilization of lymphoma and no cell death escape. Other mechanisms to upregulate this gene.
ROBO1  | healthy > cancer | Less expression in cancer samples thus downregulation: Development of invasive and metastatic tumor through loss of function of important tumor supressor gene in MCL.
SOX9   | cancer > healthy | Higher expression in cancer samples: Gain of autosomal sex-reversal in MCL.
HOXA9  | cancer > healthy | Higher expression in cancer samples: Overexpression leads to strong self-renewal and inhibition of differentiation of MCL cells.
CDH1   | healthy > cancer | Gene expression is dependent on cell cycle phase therefore can not be evaluated properly.
CDC14B | healthy > cancer | Less expression (cancer samples) and thus additional p53 deactivation and overriding cell cyle checkpoints. Enables unlimited cancer cell growth of MCL cells.
FOXC1  | cancer > healthy | Higher expression leads to abnormal growth, migration, cell proliferation and thus functions as an oncogene for MCL.
G0S2   | cancer > healthy | Upregulated gene enables MCL resistance towards treatment.
GPX3   | healthy > cancer | Downregulation leads to inactivation of an important tumor supressor gene in MCL.
LGALS3 | healthy > cancer | Downregulated therefore no drug tolerance in tested MCL samples.
MAL    | healthy > cancer | Downregulated therefore this specific MCL marker is not active.
PAX6   | cancer > healthy | Higher expression leads to apoptosis of MCL cells. Other mechanisms to downregulate this specific gene.
TFPI2  | healthy > cancer | Less expression in MCL cells and therefore important tumor supressor is disabled leading to an extracellular matrix remodeling.
THEM4  | healthy > cancer | Downregulation enables cell survival, growth, proliferation, cell migration and angiogenesis of MCL cells which are necessary for MCL survival.
TWIST1 | healthy > cancer | Less expression of TWIST1 in cancer. Other mechanisms for overexpression in cancer than methylation.

To valididate our assumptions it would be important to analyse protein concentrations in cancer cells and compare it to DNA methylation status of genes.

For better valuation of how conclusive following analysis are, it would be pretty neat, if we compare standard deviations (SD) of cancer and healthy m-values.
At first, we compared general information with following attributes shown in the plot:

* SD of cancer and healthy m-values (black points) 

```{r}
ggplot() +
  geom_point(
    mapping = aes(
      x = extended_cancer_m_values$sd_cancer_m_values,
      y = extended_healthy_m_values$sd_healthy_m_values
    ),
    na.rm = TRUE,
    alpha = 1 / 10
  ) +
  labs(x = "SD cancer m-values",
       y = "SD healthy m-values",
       title = "Comparison of SD of m-values") +
  theme_bw()
```
In general, our standard deviation is sufficient and only a few have a high standard deviation.

For further analysis, it would be proficient if cancer related genes have low standard deviations in both sample groups. The following detailed plot visualizes:

* SD of cancer and healthy m-values (black points), 
* highlights of important genes for cancer (red points), 
* labels of specific genes 

```{r}
ggplot() +
  geom_point(
    mapping = aes(
      x = extended_cancer_m_values$sd_cancer_m_values,
      y = extended_healthy_m_values$sd_healthy_m_values
    ),
    na.rm = TRUE,
    alpha = 1 / 10
  ) +
  geom_point(
    mapping = aes(
      x = extended_cancer_m_values_gene$sd_cancer_m_values,
      y = extended_healthy_m_values_gene$sd_healthy_m_values
    ),
    colour = "red",
    size = 2
  ) +
  geom_label_repel(
    aes(
      label = Important_genes,
      x = extended_cancer_m_values_gene$sd_cancer_m_values,
      y = extended_healthy_m_values_gene$sd_healthy_m_values
    ),
    data = extended_cancer_m_values_gene,
    point.padding = 0.5,
    label.size = 0.1,
    segment.colour = "cornflowerblue",
    segment.alpha = 0.9,
    max.iter = 5000
  ) +
  labs(x = "SD cancer m-values",
       y = "SD healthy m-values",
       title = "Comparison of SD of m-values") +
  xlim(-1, 2.5) +
  ylim(-1, 2.5) +
  theme_bw()
```
In general, our important genes for MCL have a low standard deviation which enhance the explanatory power of previous assumptions. A few genes tend to be more differently methylated in cancer samples than in healthy samples. For example one of this gene is SOX11.
SOX11 tends to be differently methylated in cancer samples. This observation can be made because MCL can be devided into 2 groups: SOX11 positive and SOX11 negative patients. Depending on this status, different clinical outcome can be observed.
Broadly, methylation of cancer patients have stronger discrepancies among each other because cancer in each patient differs.













$$~$$


## Part 3: Data reduction
$$~$$

The aim of this part of the project is to narrow down our data to the significant genes which account for the majority of the variation between methylation values of healthy and cancer samples. Therefore, we need to reduce high dimensional data into low dimensional data with few information loss. 
A common way to identify the direction of highest variability in the data is by conducting a principal component analysis (PCA) which we also applied to our data. Hence the covariance matrix is used to calculate the eigenvector which correspond to the largest eigenvalues (the principal components). These eigenvectors are used to reconstruct a significant fraction of the variance of the original data.
These new principal components (features) can now be used for predictive analysis and model building. A crucial benefit is, that each principal component is independent of one another.

But first we need to tweak our data. There are several +/- infinity values in our m-value dataset, with which PCA does not function. These were caused by beta values of 0 and 1. The most straightforward solution would be to approximate these values with 0.00001 and 0.99999 and calculate the m-values again.

```{r}
cancer_beta_values[cancer_beta_values == 0] <- 0.00001
cancer_beta_values[cancer_beta_values == 1] <- 0.99999
healthy_beta_values[healthy_beta_values == 0] <- 0.00001
healthy_beta_values[healthy_beta_values == 1] <- 0.99999

cancer_m_values <-
  data.frame(log2(cancer_beta_values / (1 - cancer_beta_values)))
healthy_m_values <-
  data.frame(log2(healthy_beta_values / (1 - healthy_beta_values)))
```

After that we changed the ending of the columnnames in our new m-values dataframe from .bed to .M because there are m-values in the dataframe and no beta values anymore.

```{r, echo=FALSE}
colnames(healthy_m_values) <-
  c(
    "Bcell_naive_VB_NBC_NC11_41.M",
    "Bcell_naive_VB_NBC_NC11_83.M",
    "Bcell_naive_VB_S001JP51.M",
    "Bcell_naive_VB_S00DM851.M",
    "Bcell_naive_VB_S01ECGA1.M"
  )

colnames(cancer_m_values) <- c(
  "cancer_VB_S01FE8A1.M",
  "cancer_VB_S01FF6A1.M",
  "cancer_VB_S01FH2A1.M",
  "cancer_VB_S01FJZA1.M",
  "cancer_VB_S01FKXA1.M"
)
```

Aferwards we merged the healthy and cancer m-value datasets with each other and applied the function for pca.

```{r}
m_values <- cbind(healthy_m_values, cancer_m_values)
pca_m_values <- prcomp(t(m_values))
summary(pca_m_values)

```

From the summary we can conclude that PC1 and PC2 together account for nearly 50% of the varaince in the data.
Now, we want to find out the number of PC's we want to work with using the elbow method. For that we need to calculate the proportion of variance "by hand" and plot that against the number of PC's.

```{r}
std_dev <- pca_m_values$sdev
variance <- std_dev ^ 2
prop_var <- variance / sum(variance)

plot(
  prop_var,
  main = "Variance explained by principal components",
  xlab = "Principal Components",
  ylab = "Proportion of Variance Explained",
  type = "b"
)
```

The knick in the graph is not that apparent but by eyeballing we deciding to go with the first 3 PC's.

To get a feel of how the samples were spread across PC1 and PC2 we plottet a scatterplot. Additionally we used the function `plotly()` to define each point on the plot by pointing on it.
For a better clearness we first added an extra column with the information if the sample is from a healthy person or a cancer patient to the data frame before drawing the scatterplot. This is necessary because we want to be able to colour the points in the kmeans clustering red in the case of cancer and green in case of a healthy sample.

```{r}
pcs_of_m_values <-
  data.frame(cbind(
    pca_m_values$x,
    Samples = c(
      "Healthy",
      "Healthy",
      "Healthy",
      "Healthy",
      "Healthy",
      "Cancer",
      "Cancer",
      "Cancer",
      "Cancer",
      "Cancer"
    )
  ))
p <- ggplot(pcs_of_m_values, aes(PC1, PC2, group = Samples)) +
  geom_point (aes(shape = Samples, color = Samples), size = 4) +
  theme_bw()
p <- p + scale_colour_manual(values = c("seagreen2", "indianred2"))
p <- ggplotly(p)
p
```


#In this scatter plot we can see that healthy and cancer samples are clustered into two different groups, although the individual samples group are not as tightly clustered. Healthy samples are more tightly clustered than that of cancer, because as we have seen different mutations have been associated with MCL. Another reason for long distances between samples of each cluster can be batch effects which are an unwanted source of variation in our data (Leek, 2012). Thus we need to identify which PC has the most batch effects so that we can neglcets it further down the line of our analysis. 


Now we want to perform kmeans clustering. We know that there should be two clusters (healthy/cancer) but we still wanted to additionaly prove that. So we plotted the number of clusteres against the total within-clusters sum of squares.

```{r}
wss <-  sapply(1:5, function(k) {
  kmeans(x = pca_m_values$x,
         centers = k,
         iter.max = 100)$tot.withinss
})
plot(
  1:5,
  wss,
  type = "b",
  pch = 19,
  xlab = "Number of clusters K",
  ylab = "Total within-clusters sum of squares"
)
```

The knick in the elbow is as expected at 2 clusters. 





```{r}
centers <-
  kmeans(
    x = t(pcs_of_m_values[1:10]),
    centers = 2,
    iter.max = 100
  )$centers
```


```{r}
centers <-
  data.frame(cbind(
    t(centers),
    Samples = c(
      "Healthy",
      "Healthy",
      "Healthy",
      "Healthy",
      "Healthy",
      "Cancer",
      "Cancer",
      "Cancer",
      "Cancer",
      "Cancer"
    )
  ))
```


```{r}

```

#We can see that healthy and cancer samples are clustered into two different groups, although the individual samples group are not as tightly clustered. Reason for that can be batch effects which are an unwanted source of variation in our data. Thus we need to identify which PC has the most batch effects so that we can neglect it further down the line of our analysis.











$$~$$

## Part 5: Logistic Regression

To obtain a "good" prediction system, we first had to check on correlation between our final genes. A high correlation would result in a bad regression model because of insufficient information. Correlation ranges from 0 to 1. The value 0 means no correlation, therefore, maximum information. In the contrast, the value 1 means high correlation, therefore, no information in our case. 

```{r, include = FALSE}
library(sandwich)
library(corrplot)
library(jtools)
```

```{r}
correlation_top_genes <- 
  cor(
    clustering_data
    )
corrplot(
  correlation_top_genes, 
  title = "Correlation between all samples"
  )
heatmap(
  as.matrix(t(clustering_data)), 
  main = "Heat map of samples and genes"
  )
hist(
  cor(t(clustering_data)), 
  main = "Genes correlation", xlab = "Correlation value"
  )
```

Going into details, the first plot visualizes general correlation between samples, therefore, including every gene of one sample and calculating overall correlation. The larger the point is, the greater the correlation is. The color range also indicates how much correlation there is. Blue for positive, red for negative correlation. 

The second plot shows a heat map. Through a heatmap we visualized hierarchical clustering. Data values were transformed to color scale, thus, we simultaneously identified clusters of samples and features. This process starts with hierarchical clustering of both the rows and the columns of the data matrix. Then columns/rows of the data matrix are re-ordered according to the hierarchical clustering result, putting similar observations close to each other. The blocks of ‘high’ and ‘low’ values are adjacent in the data matrix. Finally, a color scheme is applied for the visualization and the data matrix is displayed. Visualizing the data matrix in this way can help to find the variables that appear to be characteristic for each sample cluster. Thus, we identified a general trend. Almost every gene of each sample followed the same trend. This showed again the relationship of high correlation, because distances between genes were pretty small. A distance of zero indicates a perfect correlation of 1 (Distance = 1 - Correlation).

Last but not least, a complementary histogram is shown to quantify the frequency of correlation values. This plot underlines previous results of analysis.

To conclude, there is high positive correlation between our samples within a group. But between our cancer and control/healthy group we have pretty much no correlation. Hopefully, this is sufficient for our prediction model.

In the process of testing different models, we faced several problems. Firstly, we had way too many genes which were high correlated. Thus we tried to reduce the amount of genes with regard to their correlation amount. For this, we had to calculate their correlation index and set a threshold reasonably enough for further analysis.
```{r}
correlation_top_genes <- 
  cor(t(clustering_data)
      )
correlation_top_genes[upper.tri(correlation_top_genes)] <- 0
diag(correlation_top_genes) <- 0
correlation_top_genes <- 
  correlation_top_genes[ ,
    !apply(
      correlation_top_genes,
      2,function(x) {
        any(x > 0.65)
        }
      )
    ]

regression_data <- 
  clustering_data
regression_data <- 
  regression_data[
    c(
    colnames(correlation_top_genes)),  
    ]
```

Our chosen threshold was 0.65. Thus, each column (gene) which contained more than one value above this threshold was removed. Through this threshold, we obtained a reasonably mediocre regression model which was better than nothing. Our correlation relationship significantly improved  to our favour.
```{r}
correlation_top_genes <- 
  cor(
    t(regression_data)
    )
corrplot(
  correlation_top_genes
  )
```
We even have now negative correlation which maximizes our information content.
On the base of this new data frame, we try to fit a logistic regression model. We used a logistic regression model, because it manages to model a binomial outcome (0,1) with one or more explanatory variables. In our case the variables are the selected genes. It measures the relationship between the categorical dependent variable (cancer/healthy) and one or more independent variables by estimating probabilities using a logistic function, which is the cumulative logistic distribution. 
Logistic function: 
$$ p(X) = \frac{e^{β_0+β_1X}}{1+e^{β_0+β_1X}} $$

```{r}
regression_data <- 
  cbind(
    Health_status = 
          as.numeric(
            c(
              0,
              0,
              0,
              0,
              0,
              1,
              1,
              1,
              1,
              1
              )
            ), 
        as.data.frame(t(regression_data))
        )
regression_model <- 
  glm(
    formula = Health_status ~ ., 
    family = binomial(link = "logit"), 
    data = regression_data, 
    maxit = 4
    )

summary(regression_model)
summ(regression_model, confint = TRUE, digits = 3, ci.width = .9)
```
The error "did not converge" gave a first impression of our model. It is pretty bad. Non of our remaining genes are significant as further predictors. Moreover, we obtained high standard errors which underline unperfect model creation. We had to reduce iterations to 4 to prevent error message "fitted probabilities numerically 0 or 1 occurred". This indicated that both data are seperated in two sides. On the one side (cancer(1)) and on the other side (healthy(0)). Without cross validation, we tried to test this model on the same data again. 
Moreover, we calculated the value of χ-squared of our model and its p-value. Because of insufficient data it seemed that our model is even worse (p = 0.008) and did not fit our needs (if we set α-level to e.g. α = 0.01 and therfore rejecting our H0-hypothesis). 
Furthermore, when analyzing data with a logistic regression, an equivalent statistic to R-squared does not exist. However, to evaluate the goodness-of-fit of logistic models, several pseudo R-squareds have been developed. These are “pseudo” R-squareds because they look like R-squared in the sense that they are on a similar scale, ranging from 0 to 1 (though some pseudo R-squareds never achieve 0 or 1) with higher values indicating better model fit, but they cannot be interpreted as one would interpret an OLS R-squared and different pseudo R-squareds can arrive at very different values. Our logistic regression model had pretty good pseudo R-squared values (0.996 and 0.992) with two different methods. Thus, we estimated that our model has a good goodness-of-fit.

```{r}
predict(regression_model, newdata = regression_data, type = "response")
```
We were able to predict samples of patients which we used to model the regression model. Thankfully, predictions were correct and we were able to go on with cross validation.

Here can be seen the cross validation:
```{r}
control <-  
  which(regression_data$Health_status == "0")
control.train <- 
  sample(control, floor(0.75 * length(control)))
cancer <- 
  which(regression_data$Health_status == "1")
cancer.train <- 
  sample(cancer, floor(0.75 * length(cancer)))
train <- 
  c(control.train, cancer.train)
train.set <- 
  regression_data[train, ]
test.set <- 
  regression_data[-train, ]
regression_model2 <- 
  glm(formula = Health_status ~ ., family = binomial(link = "logit"), data = train.set, maxit = 4)
predict(regression_model2, newdata = regression_data, type = "response")
summ(regression_model2)
```
After splitting up the data into test and train set, we modelled a regression system based on the train set. Then we used this regression model to finally see if this model was good enough to predict the test set. In most cases it seemed like it predicted the samples correctly but there were also a few constellations in which it predicted samples wrong.
Overall χ-squared (p=0.08) and pseudo R-squared values improved slightly. This indicated an improved logistic regression model.

We now wanted to find out what the tasks of our determined genes are.
Our significant genes have following tasks:

ENSG00000262554:
Number represents Zinc finger and SCAN domain-containing protein 10.
Embryonic stem (ES) cell-specific transcription factor required to maintain ES cell pluripotency. In our case high methylation maybe activates gene expression and therefore necessary cell pluripotency is enabled (Cunningham et al., 2019).

ENSG00000273750:
Number represents HOXB13 antisense RNA 1 (non-protein coding).
A long non-coding RNA which is highly expressed in prostate, rectum, colon, and testis. No further function known (McEntyre and Birney, 2016; Cunningham et al., 2019). Methylation can not be interpreted.

ENSG00000252026:
Number represents U6 Small Nuclear 1262 (a pseudogene).
It is a part of a complex that combines with other snRNPs, unmodified pre-mRNA, and various other proteins to assemble a spliceosome, a large RNA-protein molecular complex that catalyzes the excision of introns from pre-mRNA (McEntyre and Birney, 2016; Cunningham et al., 2019). Methylation effect on this gene segment is not predictable.

ENSG00000241288:
Number represents Long Intergenic Non-Protein Coding RNA 2614.
The group of long noncoding RNAs (lncRNAs) are known to be important in cancer as they directly interact with the cell cycle, proliferation pathways and microbiome balance. Moreover, lncRNAs regulate the immune system: they do not directly encode proteins of innate or adaptive immunity, but regulate immune cell differentiation and function, such as dendritic cell activity, T cell ratio and metabolism. The result of this complex interaction is that lncRNAs regulate cancer processes through a complex multimodal system involving immunity, metabolism and infection (Denaro et al., 2019). Thus, it would be optimal for our cancer to activate this gene.

After all those analyse steps, we can conclude that the data, on which our analysis is based, is not proficient enough for our downstream analysis. We can in no way guarantee that this genes are relevant for mantle cell lymphoma.

For the sake of completeness, the following logistic regression model included all of the genes after evaluating through t-test and p-adjustment.
```{r}
regression_data2 <- cbind(
  Health_status = as.numeric(
    c(
  0,
  0,
  0,
  0,
  0,
  1,
  1,
  1,
  1,
  1
)), as.data.frame(t(clustering_data)))
regression_model3 <- 
  glm(
    formula = Health_status ~ ., 
    family = binomial(link = "logit"), 
    data = regression_data2
    )

summary(regression_model3)
```
As previous mentioned, this model is as bad as it possibly can be. We faced with alot of NA values based on highly correlated data on the one and on the other hand we had to deal with values of 1. We rejected this model and did some extra work to create a better model shown above.

# References

Sharp, A.J., Stathaki, E., Migliavacca, E., Brahmachary, M., Montgomery, S.B., Dupre, Y., and Antonarakis, S.E. (2011). DNA methylation profiles of human active and inactive X chromosomes. Genome Res 21, 1592-1600.

n, B.C., Choi, Y.D., Oh, I.J., Kim, J.H., Park, J.I., and Lee, S.W. (2018). GPx3-mediated redox signaling arrests the cell cycle and acts as a tumor suppressor in lung cancer cell lines. PLoS One 13, e0204170.

Brumatti, G., Salmanidis, M., Kok, C.H., Bilardi, R.A., Sandow, J.J., Silke, N., Mason, K., Visser, J., Jabbour, A.M., Glaser, S.P., et al. (2013). HoxA9 regulated Bcl-2 expression mediates survival of myeloid progenitors and the severity of HoxA9-dependent leukemia. Oncotarget 4, 1933-1947.

Cunningham, F., Achuthan, P., Akanni, W., Allen, J., Amode, M.R., Armean, I.M., Bennett, R., Bhai, J., Billis, K., Boddu, S., et al. (2019). Ensembl 2019. Nucleic Acids Res 47, D745-D751.

Denaro, N., Merlano, M.C., and Lo Nigro, C. (2019). Long noncoding RNAs as regulators of cancer immunity. Mol Oncol 13, 61-73.

Dong, Y., Tan, Q., Tao, L., Pan, X., Pang, L., Liang, W., Liu, W., Zhang, W., Li, F., and Jia, W. (2015). Hypermethylation of TFPI2 correlates with cervical cancer incidence in the Uygur and Han populations of Xinjiang, China. Int J Clin Exp Pathol 8, 1844-1854.

Enjuanes, A., Albero, R., Clot, G., Navarro, A., Bea, S., Pinyol, M., Martin-Subero, J.I., Klapper, W., Staudt, L.M., Jaffe, E.S., et al. (2013). Genome-wide methylation analyses identify a subset of mantle cell lymphoma with a high number of methylated CpGs and aggressive clinicopathological features. Int J Cancer 133, 2852-2863.

Enjuanes, A., Fernandez, V., Hernandez, L., Navarro, A., Bea, S., Pinyol, M., Lopez-Guillermo, A., Rosenwald, A., Ott, G., Campo, E., et al. (2011). Identification of methylated genes associated with aggressive clinicopathological features in mantle cell lymphoma. PLoS One 6, e19736.

Fei, F., Joo, E.J., Tarighat, S.S., Schiffer, I., Paz, H., Fabbri, M., Abdel-Azim, H., Groffen, J., and Heisterkamp, N. (2015). B-cell precursor acute lymphoblastic leukemia and stromal cells communicate through Galectin-3. Oncotarget 6, 11378-11394.

Gort, E.H., Suijkerbuijk, K.P., Roothaan, S.M., Raman, V., Vooijs, M., van der Wall, E., and van Diest, P.J. (2008). Methylation of the TWIST1 promoter, TWIST1 mRNA levels, and immunohistochemical expression of TWIST1 in breast cancer. Cancer Epidemiol Biomarkers Prev 17, 3325-3330.

Hu, H., Chen, X., Wang, C., Jiang, Y., Li, J., Ying, X., Yang, Y., Li, B., Zhou, C., Zhong, J., et al. (2017). The role of TFPI2 hypermethylation in the detection of gastric and colorectal cancer. Oncotarget 8, 84054-84065.

Imamura, F., and Greer, C.A. (2013). Pax6 regulates Tbr1 and Tbr2 expressions in olfactory bulb mitral cells. Mol Cell Neurosci 54, 58-70.

Klauck, P.J., Bagby, S.M., Capasso, A., Bradshaw-Pierce, E.L., Selby, H.M., Spreafico, A., Tentler, J.J., Tan, A.C., Kim, J., Arcaroli, J.J., et al. (2018). Antitumor activity of the polo-like kinase inhibitor, TAK-960, against preclinical models of colorectal cancer. BMC Cancer 18, 136.

Kuo, P.Y., Jatiani, S.S., Rahman, A.H., Edwards, D., Jiang, Z., Ahr, K., Perumal, D., Leshchenko, V.V., Brody, J., Shaknovich, R., et al. (2018). SOX11 augments BCR signaling to drive MCL-like tumor development. Blood 131, 2247-2255.

Li, Y., Han, L., Huang, C., Dai, W., Tian, G., Huang, F., Li, J., Liu, J., Wang, Q., and Zhou, Z. (2018). New Contributions to Asarum Powder on Immunology Related Toxicity Effects in Lung. Evid Based Complement Alternat Med 2018, 1054032.

McEntyre, J., and Birney, E. (2016). The EMBL-EBI channel. F1000Res 5, 52.

Meng, B., Wang, Y., and Li, B. (2014). Suppression of PAX6 promotes cell proliferation and inhibits apoptosis in human retinoblastoma cells. Int J Mol Med 34, 399-408.

Mohanty, S., Mohanty, A., Sandoval, N., Tran, T., Bedell, V., Wu, J., Scuto, A., Murata-Collins, J., Weisenburger, D.D., and Ngo, V.N. (2017). Cyclin D1 depletion induces DNA damage in mantle cell lymphoma lines. Leuk Lymphoma 58, 676-688.

Ondracka, A., Robbins, J.A., and Cross, F.R. (2016). An APC/C-Cdh1 Biosensor Reveals the Dynamics of Cdh1 Inactivation at the G1/S Transition. PLoS One 11, e0159166.

Parray, A., Siddique, H.R., Kuriger, J.K., Mishra, S.K., Rhim, J.S., Nelson, H.H., Aburatani, H., Konety, B.R., Koochekpour, S., and Saleem, M. (2014). ROBO1, a tumor suppressor and critical molecular barrier for localized tumor cells to acquire invasive phenotype: study in African-American and Caucasian prostate cancer models. Int J Cancer 135, 2493-2506.

Puttabyatappa, M., Al-Alem, L.F., Zakerkish, F., Rosewell, K.L., Brannstrom, M., and Curry, T.E., Jr. (2017). Induction of Tissue Factor Pathway Inhibitor 2 by hCG Regulates Periovulatory Gene Expression and Plasmin Activity. Endocrinology 158, 109-120.

Rajapakse, V.N., Luna, A., Yamade, M., Loman, L., Varma, S., Sunshine, M., Iorio, F., Sousa, F.G., Elloumi, F., Aladjem, M.I., et al. (2018). CellMinerCDB for Integrative Cross-Database Genomics and Pharmacogenomics Analyses of Cancer Cell Lines. iScience 10, 247-264.

van der Post, R.S., Vogelaar, I.P., Carneiro, F., Guilford, P., Huntsman, D., Hoogerbrugge, N., Caldas, C., Schreiber, K.E., Hardwick, R.H., Ausems, M.G., et al. (2015). Hereditary diffuse gastric cancer: updated clinical guidelines with an emphasis on germline CDH1 mutation carriers. J Med Genet 52, 361-374.

Vogel, C.F., Li, W., Sciullo, E., Newman, J., Hammock, B., Reader, J.R., Tuscano, J., and Matsumura, F. (2007). Pathogenesis of aryl hydrocarbon receptor-mediated development of lymphoma is associated with increased cyclooxygenase-2 expression. Am J Pathol 171, 1538-1548.

Wang, Q., He, Y., Shen, Y., Zhang, Q., Chen, D., Zuo, C., Qin, J., Wang, H., Wang, J., and Yu, Y. (2014). Vitamin D inhibits COX-2 expression and inflammatory response by targeting thioesterase superfamily member 4. J Biol Chem 289, 11681-11694.

Wang, Y., Hou, Y., Zhang, W., Alvarez, A.A., Bai, Y., Hu, B., Cheng, S.Y., Yang, K., Li, Y., and Feng, H. (2019). Lipolytic inhibitor G0S2 modulates glioma stem-like cell radiation response. J Exp Clin Cancer Res 38, 147.

Wasik, A.M., Lord, M., Wang, X., Zong, F., Andersson, P., Kimby, E., Christensson, B., Karimi, M., and Sander, B. (2013). SOXC transcription factors in mantle cell lymphoma: the role of promoter methylation in SOX11 expression. Sci Rep 3, 1400.

Welch, C., Santra, M.K., El-Assaad, W., Zhu, X., Huber, W.E., Keys, R.A., Teodoro, J.G., and Green, M.R. (2009). Identification of a protein, G0S2, that lacks Bcl-2 homology domains and interacts with and antagonizes Bcl-2. Cancer Res 69, 6782-6789.

Leek, J.T., Johnson, W.E., Parker, H.S., Jaffe, A.E., and Storey, J.D. (2012). The sva package for removing batch effects and other unwanted variation in high-throughput experiments. BIOINFORMATICS 28, 2 S.

