---
title: "Are differentially methylated regions within genes associated with mantle cell lymphoma?"
author: "Pascal Lafrenz, Mari Hambardzumyan, Lea Herzel, Franziska Lam"
output:
  pdf_document: default
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---

The alterations in cancer cell DNA methylation pattern have long been detected (Vogelstein and Feinberg, 1983). In many cancer types methylation level has also been shown to have effects on gene deregulation,  (Melnick, 2013). But not much research has been done associated with methylation in *mantle cell lymphoma* (MCL) cells. Therefore in the following project we want to address the following question.

# Research question

Are differentially methylated regions associated with mantle cell lymphoma?

For that purpose we received a data set and an annotation file. Below is a short summary of important key points of the data set.

# Data summary

* Bisulfite-Seq data for 10 patients/samples (5 male, 5 female)
* 5 healthy patients, 5 with MCL
* Extracted DNA is from B-cells
* 56175 genes across all samples
* coverage and beta-values for each gene
* Length of each gene

Researching on MCL we found a range of genes which have been linked with this specific cancer, some of which include SOX11, p53, CCND1 etc (Queirós et al., 2016).

To address the research question we broke down the project into 5 parts.

# Project steps



|1. Data processing  | 2. Data normalization and visualization | 3. Data reduction  | 4. Regression and interpretation    | 5. Annotate and interpret results
------------- | ------------- | ------------- | ------------- | -------------
Remove x and y chromosome genes  | Transform beta-values into homoscedastic M-values  | Dimensionality reduction, e.g PCA  | Apply t-test        |  
Split data in 4 data frames (cancer vs. control, beta-values and coverage)  | Add a column for mean, median and SD for better comparisons  | Apply clustering method and visualize, e.g k-means         | Apply logistic regression        | 
Set a threshold and examine variance for reliable coverage, replace unreliable beta-values with NA | Visualization through plots | 
Separately inspect genes of interest for MCL | Identify batch effect magnitude in the data set
Set a threshold and remove genes under/above threshold, imputate remaining NA's | 

$$~$$

## Part 1: Data processing

For now, we want to single out the gene data frame of the patients to inspect it separately. We will examine only the gene data frame and neglect the data from tiling, promoters and cpg-islands because they are irrelevant to answer our research question.

```{r}
Samples <- readRDS(file = "Mantle-Bcell_list.RDS.gz")
Gene_data_frame <- Samples$genes
dim(Gene_data_frame)
```

There are much information from 56175 genes across all 10 patients. Mostly we are interested in the coverage and the beta values. After the entire data clean-up we still want to retain around 90% of our beta value information (~500000 genes).

We are starting the data clean-up with the sex chromosomes. In our case we have different numbers of female/male patients in healthy/cancer sample groups. In female X chromosome inactivation leads to the silencing of the majority of genes on one X, this would lead to significant differences in the methylation profiles between female/male patients (Sharp et al., 2011). Therefore, we remove gene data of sex chromosomes. 

```{r}
Gene_data_frame_x_y <- 
  Gene_data_frame[-which(Gene_data_frame$Chromosome == "chrX"),]
Gene_data_frame_x_y <- 
  Gene_data_frame_x_y[-which(Gene_data_frame_x_y$Chromosome == "chrY"),]
```

Now we want to go through the coverage values separately to remove any unreliable values, outliers and possible systematic errors. First we separated the coverage values of healthy and cancer patients into 2 data frames. We did the same separation into 2 data frames with the beta-values because we will need these later. All the  other information will be neglectd till the interpretation. 

```{r}
healthy_coverage <- Gene_data_frame_x_y[, 21:25]
cancer_coverage <- Gene_data_frame_x_y[, 26:30]
healthy_beta_values <- Gene_data_frame_x_y[, 11:15]
cancer_beta_values <- Gene_data_frame_x_y[, 16:20]
```

Then we created a histogram with mean coverage values for each gene on a log scale and added 10 quantiles from 10 % to 100 %.

```{r}
mean_cancer_coverage <- rowMeans(cancer_coverage)
hist(
  log10(mean_cancer_coverage),
  breaks = "fd",
  main = "Cancer coverage: Mean frequency",
  xlab = "Common logarithm of coverages",
  col = "indianred1",
  border = "gray20"
)
abline(v = log10(quantile(
  mean_cancer_coverage,
  probs = seq(0, 1, 0.1),
  na.rm = TRUE
)),
col = "black",
lty = 5,
lwd = 1)


mean_healthy_coverage <- rowMeans(healthy_coverage)
hist(
  log10(mean_healthy_coverage),
  breaks = "fd",
  main = "Healthy coverage: Mean frequency",
  xlab = "Common logarithm of coverages",
    col = "seagreen2",
    border = "gray20"
)
abline(v = log10(quantile(
  mean_healthy_coverage,
  probs = seq(0, 1, 0.1),
  na.rm = TRUE
)),
col = "black",
lty = 5,
lwd = 1)
```

To check how widely spread are the coverages across the patients, we calculated the standard deviations (SD) of each gene across the patients. We plotted the SD-values for cancer and healthy samples on a histogram with a log scale.

```{r}
sd_cancer_coverage <- apply(cancer_coverage, 1, sd)
hist(log10(sd_cancer_coverage),
     breaks = "fd",
     main = "Cancer coverage: SD frequency",
     xlab = "Common logarithm of coverages",
     col = "indianred1",
     border = "gray20"
)
sd_healthy_coverage <- apply(healthy_coverage, 1, sd)
hist(log10(sd_healthy_coverage),
     breaks = "fd",
     main = "Healthy coverage: SD frequency",
     xlab = "Common logarithm of coverages",
     col = "seagreen2",
     border = "gray20"
)
```

By setting the lower coverage threshold to 5 % and the upper to 99.9 %, we fulfilled our initial requirement of retaining around 90 % of the information. 

```{r}
threshold <-
  quantile(mean_cancer_coverage,
           probs = seq(0.05, 0.05),
           na.rm = TRUE)
threshold2 <-
  quantile(mean_cancer_coverage,
           probs = seq(0.999, 0.999),
           na.rm = TRUE)

threshold3 <-
  quantile(mean_healthy_coverage,
           probs = seq(0.05, 0.05),
           na.rm = TRUE)
threshold4 <-
  quantile(mean_healthy_coverage,
           probs = seq(0.999, 0.999),
           na.rm = TRUE)
```

Beta-values with very low or extreme high coverages are not reliable. Therefore with the help of the predefined coverage threshold for each patient we will set the beta values of the genes which didn't fulfill the coverage requirement to NA.

```{r}
for (i in 1:nrow(cancer_coverage)) {
  for (j in 1:ncol(cancer_coverage)) {
    if (cancer_coverage[i, j] <= threshold) {
      cancer_coverage[i, j] <- 0
    }
    
    if (cancer_coverage[i, j] >= threshold2) {
      cancer_coverage[i, j] <- 0
    }
    
    if (cancer_coverage[i, j] == 0) {
      cancer_coverage[i, j] <- NA
      cancer_beta_values[i, j] <- NA
      
    }
  }
}

for (i in 1:nrow(healthy_coverage)) {
  for (j in 1:ncol(healthy_coverage)) {
    if (healthy_coverage[i, j] <= threshold3) {
      healthy_coverage[i, j] <- 0
    }
    
    if (healthy_coverage[i, j] >= threshold4) {
      healthy_coverage[i, j] <- 0
    }
    
    if (healthy_coverage[i, j] == 0) {
      healthy_coverage[i, j] <- NA
      healthy_beta_values[i, j] <- NA
      
    }
  }
}
```

To proof the success of our data cleanup we plotted the cancer and healthy coverage histograms again. In both cases it is seen that the shapes of the bars are still the same, but the low coverage values are deleted.

```{r}
mean_cancer_coverage <- rowMeans(cancer_coverage)
hist(
  log10(mean_cancer_coverage),
  breaks = "fd",
  main = "Cancer coverage: Mean frequency",
  xlab = "Common logarithm of coverages",
  col = "indianred1",
  border = "gray20"
)
abline(v = log10(quantile(
  mean_cancer_coverage,
  probs = seq(0, 1, 0.1),
  na.rm = TRUE
)),
col = "black",
lty = 5,
lwd = 1)


mean_healthy_coverage <- rowMeans(healthy_coverage)
hist(
  log10(mean_healthy_coverage),
  breaks = "fd",
  main = "Healthy coverage: Mean frequency",
  xlab = "Common logarithm of coverages",
  col = "seagreen2",
  border = "gray20"
)
abline(v = log10(quantile(
  mean_healthy_coverage,
  probs = seq(0, 1, 0.1),
  na.rm = TRUE
)),
col = "black",
lty = 5,
lwd = 1)

```

Now we needed to deal with NA's in the beta-values, because they represent no information and they could cause errors in our further analysis. We have many NA's but we cannot remove every gene with a NA because this would lead to a loss of too much information.

```{r}
sum(is.na(cancer_beta_values))
sum(is.na(healthy_beta_values))
```

We decided that we want to remove each gene with 3 or more NA's in 'healthy_beta_values' or 'cancer_beta_values', because if there are not enough values our results will be not meaningful. To do so we added two new columns to each data frame. One column is for the number of NA's per gene in healthy samples and one for the number of NA's per gene in cancer samples. We need both columns in both data frames because it makes it easier to remove the rows with too much NA's later.

```{r}
cancer_beta_values$Number_of_NA_cancer <-
  rowSums(is.na(cancer_beta_values))
cancer_beta_values$Number_of_NA_healthy <-
  rowSums(is.na(healthy_beta_values))

healthy_beta_values$Number_of_NA_healthy <-
  rowSums(is.na(healthy_beta_values))
healthy_beta_values$Number_of_NA_cancer <-
  rowSums(is.na(cancer_beta_values))
```

To visualize the number of NA's in the beta-values we plotted differnt histograms.

```{r}
hist(
  cancer_beta_values$Number_of_NA_cancer,
  main =  "NA's per Gene in MCL Samples",
  breaks = 5,
  xlab = "Number of NA's",
  ylab = "Number of Genes",
  col = "indianred1",
  border = "gray20"
)
hist(
  cancer_beta_values$Number_of_NA_cancer,
  main =  "NA's per Gene in MCL Samples (zoomed in)",
  breaks = 4,
  xlim = c(1,5),
  ylim = c(0,2500),
  xlab = "Number of NA's",
  ylab = "Number of Genes",
  col = "indianred1",
  border = "gray20"
)

hist(
  healthy_beta_values$Number_of_NA_healthy,
  main =  "NA's per Gene in Healthy Samples",
  breaks = 5,
  xlab = "Number of NA's",
  ylab = "Number of Genes",
  col = "seagreen2",
  border = "gray20"
)
hist(
  healthy_beta_values$Number_of_NA_healthy,
  main =  "NA's per Gene in Healthy Samples (zoomed in)",
  breaks = 4,
  xlim = c(1,5),
  ylim = c(0,2500),
  xlab = "Number of NA's",
  ylab = "Number of Genes",
  col = "seagreen2",
  border = "gray20"
)
```

The first histogram of cancer/healthy is to get an overview over all numbers of missing values per gene. Most genes in 'cancer_beta_values' and 'healthy_beta_values' have no NA's. That is good, because it means we have much information in our analysis and our results will be more meaningful.
We wanted to see the differences between 2, 3, 4 and 5 NA's per gene better. To do so we reduced the range of the axis. Now it is visible that in 'healty_beta_values' a little amount of genes have 5 NA's. In comparison to that 'cancer_beta_values' have more genes with 5 NA's but less genes with 2, 3 or 4 NA's.


Then we removed the genes with 3 or more missing values. Here it was important to remove the genes in both data frames, because a gene in only one data frame without values to compare is meaningless. After the deletion we checked if the rownames are the same in both data frames and it was the case. This is important because otherwise the wrong rows would be compared later and the results would be completely wrong. 

```{r}
cancer_beta_values <-
  cancer_beta_values[-which(
    cancer_beta_values$Number_of_NA_cancer >= 3 |
      cancer_beta_values$Number_of_NA_healthy >= 3
  ),]
healthy_beta_values <-
  healthy_beta_values[-which(
    healthy_beta_values$Number_of_NA_cancer >= 3 |
      healthy_beta_values$Number_of_NA_healthy >= 3
  ),]
sum(rownames(healthy_beta_values) != rownames(cancer_beta_values))
```

We removed the columns with the NA's-values because we did not need them anymore. 

```{r}
cancer_beta_values <-
  cancer_beta_values[,-which(
    colnames(cancer_beta_values) %in% c("Number_of_NA_cancer", "Number_of_NA_healthy")
  )]
healthy_beta_values <-
  healthy_beta_values[,-which(
    colnames(healthy_beta_values)  %in%  c("Number_of_NA_cancer", "Number_of_NA_healthy")
  )]

```

We can not keep the remaining NA's in the beta-values because this will cause errors in our downstream analysis. But we also did not want to delete the genes, so we decided to replace the NA's with the mean value of the respective gene.

?genauer auf transposed data frames eingehen?

```{r}
transposed_cancer_beta_values <- t(cancer_beta_values)
transposed_healthy_beta_values <- t(healthy_beta_values)

for (i in 1:ncol(transposed_cancer_beta_values)) {
  transposed_cancer_beta_values[is.na(transposed_cancer_beta_values[, i]), i] <-
    mean(transposed_cancer_beta_values[, i], na.rm = TRUE)
}
for (i in 1:ncol(transposed_healthy_beta_values)) {
  transposed_healthy_beta_values[is.na(transposed_healthy_beta_values[, i]), i] <-
    mean(transposed_healthy_beta_values[, i], na.rm = TRUE)
}

cancer_beta_values <- data.frame(t(transposed_cancer_beta_values))
healthy_beta_values <- data.frame(t(transposed_healthy_beta_values))
```

We proofed that there are no more NA's left in the cancer and healthy data frames.

```{r}
sum(is.na(transposed_cancer_beta_values))
sum(is.na(transposed_healthy_beta_values))
```

There are some important genes for MCL which have ofen a different methylation pattern in comparison to heatlhy b-cells. These genes are:

Gene     |Ensembl
---------|----------------
SOX11	   |ENSG00000176887	
NR2F2	   |ENSG00000185551	
p53  	   |ENSG00000141510	
AHR  	   |ENSG00000106546		
ROBO1	   |ENSG00000169855		
SOX9 	   |ENSG00000125398		
HOXA9	   |ENSG00000078399		
CDH1 	   |ENSG00000039068		
CDC14B	 |ENSG00000081377		
FOXC1	   |ENSG00000054598	
G0S2	   |ENSG00000123689		
GPX3	   |ENSG00000211445		
LGALS3	 |ENSG00000131981		
MAL  	   |ENSG00000172005		
NPTX2	   |ENSG00000106236		 
PAX6	   |ENSG00000007372		
TFPI2	   |ENSG00000105825		
THEM4	   |ENSG00000159445		
TWIST1	 |ENSG00000122691		

To be sure we did not removed these genes in our data processing we tested if they are still all in our data frames. Because each of the 2 data frames contains the same genes it is enough to check if they are part of one of them.

```{r}
important_genes <-
  data.frame(
    c(
      "ENSG00000176887",
      "ENSG00000185551",
      "ENSG00000141510",
      "ENSG00000110092",
      "ENSG00000106546",
      "ENSG00000169855",
      "ENSG00000125398",
      "ENSG00000078399",
      "ENSG00000039068",
      "ENSG00000081377",
      "ENSG00000054598",
      "ENSG00000123689",
      "ENSG00000211445",
      "ENSG00000131981",
      "ENSG00000172005",
      "ENSG00000106236",
      "ENSG00000007372",
      "ENSG00000105825",
      "ENSG00000159445",
      "ENSG00000122691"
    )
  )
important_genes_in_data_set <- data.frame()
for (i in 1:nrow(important_genes)) {
  important_genes_in_data_set[i, ] <-
    cancer_beta_values[which(row.names(cancer_beta_values) == important_genes[i, ]), ]
}
sum(is.na(important_genes_in_data_set))
```































# References

Sharp, A.J., Stathaki, E., Migliavacca, E., Brahmachary, M., Montgomery, S.B., Dupre, Y., and Antonarakis, S.E. (2011). DNA methylation profiles of human active and inactive X chromosomes. Genome Res 21, 1592-1600.